---
title: "Quantitative variables"
subtitle: "Independent samples"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)

library(pander)
```


## Birth weights


```{r }
library(tidyverse)
d <- MASS::birthwt %>% 
  mutate(race = factor(race, levels = 1:3, labels = c("white", "black", "other")),
         bw = factor(low, levels = c(1,0), labels = c("Low", "Normal")),
         smoke = factor(smoke, levels = c(1,0), labels = c("smoker", "non_smoker")),
         lwt_kg = 0.453592 * lwt)  # mother's weight: convert pounds to kg
head(d, 3) 
```


We want to compare the weight at birth (`bwt`) in smokers and non-smokers (independent distributions)

## Tests to compare two independent distributions 

If distributions are about normal$^1$: $\quad$ Parametric tests

- If *homoscedastic*: $\quad$ two-sample *t*-test (or unpaired *t*-test)

- If *heteroscedastic*: $\quad$ Welch test (a *t*-test with Satterthwaite correction)

Otherwise: $\quad$  Distribution-free ("non-parametric") tests 

- Wilcoxon rank sum test $\quad \equiv \quad$ Mann-Whitney test

</br>

$^1$ In practice, parametric tests work well if:  
    - both distributions are (approximately) symmetrical, and  
    - have no *influential* outliers

which can be inspected in a boxplot

## Boxplot (with means overlaid)

```{r fig.height=2, echo=FALSE}
library(ggformula)
gf_boxplot(bwt ~ smoke, data = d) %>% 
  gf_summary(fun = mean, geom = "point", color = "red") %>%  
  gf_refine(coord_flip())

```

In this case:

- distributions are (approximately) symmetrical

- a single outlier in smokers, but not influential (mean = median) 

- *heteroscedasticity* (different variances) ?

Parametric test is OK (Welch test, safer)

## Welch test (and CI for means difference)

Compares the group **means**, assuming *normal distribution* in both groups) 

```{r}
t.test(bwt ~ smoke, data = d)                        # Welch test
```

## Unpaired *t*-test (and CI for difference of means)

Compares the group **means**, assuming *normal distribution* in both groups, and homoscedasticity.

```{r}
t.test(bwt ~ smoke, data = d, var.equal = TRUE)      # t-test
```




## Why test equallity of **means**?

<center>
```{r echo=FALSE}
z <- seq(-5, 25, length = 1000)
fz <- dnorm(z, 0, 1)
fx <- dnorm(z, 5, 1)
fy <- dnorm(z, 20, 1)
plot(z, fz, type='l', lwd=2, ylim=c(0,.5), main="Same variance (1), different means: 0, 5 and 20", ylab="", xlab="",  yaxs="i")
lines(z, fx, lwd=2, col="red")
lines(z, fy, lwd=2, col="blue")
lines(z, fz, lwd=2, col="black")
```
</center>

If two distributions are **normal** and have **equal variance**, then either:  
- they have different mean, or  
- they are identical


## Quantile-Quantile (QQ) plots, normal plots

In case you want to assess normality better than with a boxplot (not really needed), use this graphic

```{r echo=FALSE, fig.height=3.5, fig.width=7, fig.align='center'}
gf_qq( ~ bwt | smoke, data = d) %>% gf_qqline(col="red")
```


- normality tests (e.g., Shapiro-Wilks) is NOT advised:  
normality can never be shown; rejection depends on power


## When conditions for Welch/*t*-test are not met  

Suppose we want to compare the mother's weight in smokers and non-smokers

```{r fig.height=2, echo=FALSE}
gf_boxplot(lwt_kg ~ smoke, data = d, ylab = "mother's weight (kg)") %>% 
  gf_summary(fun.y=mean, geom="point", color="red") %>% 
  gf_refine(coord_flip())
```

In this case:

- distributions are clearly assymmetrical

- influential outliers in both groups (mean shifted wrt median) 

Distribution-free test (Wilcoxon RST)


## Wilcoxon rank sum test


```{r}
wilcox.test(lwt_kg ~ smoke, data = d)
```


## Overlaying means in boxplots

```{r fig.height=2.5, fig.align='center'}
library(ggformula)
                                                  
gf_boxplot(lwt_kg ~ smoke, data = d, ylab = "mother's weight (kg)") %>% 
  gf_summary(fun = mean, geom = "point", color = "red") %>% 
  gf_refine(coord_flip())

```


## Exercise

Dataset `ais` in package `DAAG`: hematocrite (`hc`) and `sex` of 202 athletes.

```{r}
library(DAAG)
head(ais, 4)
```

1. Is the hematocrite different in males and females? Inspect boxplots to decide if a parametric test is appropriate.

2. Give an estimate of the effect of sex on the hematocrite.

3. Would a distribution-free test produce a different result?


## Exercise (cont.)

Run the following script to read `hta.xls` data keeping only cases with at least one BP measurement and complete data for sex and total cholesterol:

```{r eval = FALSE}
d <- rio::import("data/hta.xlsx") %>% 
  filter(!is.na(sbp_v1)) %>% 
  select(sex, total_c) %>% 
  mutate(sex = factor(sex, 1:2, c("male", "female"))) %>% 
  na.omit()
```

\

1. Is there any difference in total cholesterol levels of males and females?
