---
title: "Analysis of Variance (ANOVA)"
subtitle: "Independent samples"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)
```

## Birth weights

```{r}
library(tidyverse)

d <- MASS::birthwt %>% 
  mutate(race = factor(race, levels = 1:3, labels = c("white", "black", "other")),
         low = factor(low, levels = c(1,0), labels = c("Low", "Normal")),
         smoke = factor(smoke, levels = c(1,0), labels = c("smoker", "non_smoker")),
         lwt_kg = 0.453592 * lwt)  # mother's weight: convert pounds to kg)

head(d, 4)        

library(mosaic)
library(ggformula)
```

## Birth weight related to race?

```{r fig.height=2.5}
tally(~race, data = d)

gf_boxplot(bwt ~ race, data = d) %>% 
  gf_summary(fun = mean, geom = "point", color = "red") %>% 
  gf_refine(coord_flip())

```

## ANOVA 

```{r}
result <- aov(bwt ~ race, data = d)      # sums of squares
anova(result)                            # ANOVA table

```

\

Evidence that means of `bwt` are not all equal (p = 0.008).


## Normality assessment

```{r fig.align='center'}
d$residuals <- result$residuals                         # save residuals to dataframe d
gf_qq(~residuals, data = d) %>%  gf_qqline(col="red")   # QQ-plot 

```


## Normality and homocedasticity tests

```{r}
shapiro.test(d$residuals)                               # normality test
bartlett.test(bwt ~ race, data = d)                     # homocedasticity
```

\

Tests not recommended.


## Pairwise comparisons (*a posteriori*): Tukey {.smaller}

To control the family-wise type I error rate (conservative).

```{r}
library(lsmeans)
lsmeans(result, pairwise~race)

```


## Pairwise comparisons (*a posteriori*): FDR {.smaller}

False Discovery Rate (FDR), is less conservative (more powerful).

```{r}
lsmeans::lsmeans(result, pairwise ~ race, adjust="fdr")
```


Multiple comparisons is a dificult topic. See [here](https://benwhalley.github.io/just-enough-r/multiple-comparisons.html) for more.

## Mother's weight homogeneous accross races? 

```{r fig.height = 3, fig.align='center'}
gf_boxplot(lwt_kg ~ race, data = d) %>% 
  gf_summary(fun = mean, geom = "point", color = "red") %>% 
  gf_refine(coord_flip())

```


## Kruskal-Wallis test (distribution-free)

```{r}
kruskal.test(lwt_kg ~ race, data = d) 

```

\

Evidence that distributions of `bwt` are not homogeneous (p < 0.001).


## Exercise

With the birthweight data, compute a new variable grouping the number of physician visits during the first trimester (`ftv`) as: "none", "one", "2++".  
Using this variable, try to answer these questions:

1. Are there differences in the birth weight (`bwt`) across these three groups?

1. Are there differences in the weight of mothers (`lwt_kg`) across these three groups?

In each case, decide the type of analysis by inspection of boxplots.
