---
title: "ROC curves and AUC"
subtitle: "with package `pROC`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)


```

## Prognosis of subarachnoid hemorrhage

Dataframe `aSAH` in package `pROC` shows data from [this study](https://www.ncbi.nlm.nih.gov/pubmed/19760205)

```{r}
library(pROC); data(aSAH); head(aSAH)
```

Brain injury biomarkers `ndka` and `s100b`, assessed at ICU admission, tend to be elevated in patients with poor `outcome` at 6 months.

Want to analize their accuracy to predict outcome.


## Data visualization 

```{r fig.height=4, fig.align='center'}
library(dplyr); library(ggformula); library(patchwork)
p1 <- aSAH %>% gf_boxplot(ndka ~ outcome) %>% gf_refine(coord_transform(y = "log10"))
p2 <- aSAH %>% gf_boxplot(s100b ~ outcome) %>% gf_refine(coord_transform(y = "log10"))
p1 + p2
  
```


## ndka (ROC curve, AUC and 95% CI)

<div class="columns-2">

```{r fig.height=4, fig.width=4, fig.align='left'}
roc1 <- roc(outcome ~ ndka, data = aSAH)
plot.roc(roc1, col="blue", 
         print.auc = TRUE)

auc(roc1)
ci.auc(roc1)
set.seed(1)
ci.auc(roc1, method = "bootstrap")
```


## s100b (ROC curve, AUC and 95% CI)

<div class="columns-2">

```{r fig.height=4, fig.width=4, fig.align='left'}
roc2 <- roc(outcome ~ s100b, data = aSAH)
plot.roc(roc2, col="darkgreen", 
         print.auc=TRUE)

auc(roc2)
ci.auc(roc2)
set.seed(1)
ci.auc(roc2, method = "bootstrap")

```

For more plotting options, see the help `?plot.roc` 


## Comparing AUC's of ROC curves: DeLong's test

```{r}
roc.test(roc1, roc2)
```


No evidence of different accuracy (p = 0.164).

Uses normal approximation.


## Comparing AUC's of ROC curves: bootstrap

```{r}
set.seed(1)                                   # for reproducibility
roc.test(roc1, roc2, method = "bootstrap")
```

\

No evidence of different accuracy (p = 0.160).

## Plotting both ROC curves

```{r fig.height=4, fig.width=5, fig.align='center'}
plot.roc(roc1, col="blue", print.auc=TRUE, 
         main = "ROC curves for ndka (blue) & s100b (green)")
plot.roc(roc2, col="darkgreen", print.auc=TRUE, 
         add = TRUE, print.auc.x = 0.5, print.auc.y = 0.4)
```


## "Best" threshold (max. Youden index)

<div class="columns-2">

```{r fig.height=4, fig.width=4, fig.align='left'}
plot.roc(roc2, col="darkgreen", 
         print.auc=TRUE, 
         print.thres = "best")
aSAH %>% gf_boxplot(s100b ~ outcome) %>% 
  gf_refine(coord_transform(y = "log10")) %>% 
  gf_hline(yintercept = 0.205, lty=2, col = "red")

```


## Plotting ROC curves with ggplot2 graphic system

The `plot.roc()` function uses base R graphics 

To produce a ggplot2 graphic, use `ggroc()`

```{r fig.align='center', fig.height=3}
plot1 <- ggroc(roc1, col = "blue")
plot2 <- ggroc(list(ndka=roc1, s100b=roc2))
plot1 + plot2
```


## A function for plotting 95% CI bands {.smaller}

Adapted from [here](https://rpubs.com/vguillem/465086).

```{r}
roc_with_ci <- function(obj) {
  ciobj <- ci.se(obj, specificities = seq(0, 1, l = 25))
  dat.ci <- data.frame(x = as.numeric(rownames(ciobj)),
                       lower = ciobj[, 1],
                       upper = ciobj[, 3])
  
  ggroc(obj, col = "blue") +
    theme_minimal() +
    geom_abline(
      slope = 1,
      intercept = 1,
      linetype = "dashed",
      alpha = 0.7,
      color = "darkgrey"
    ) + coord_equal() +
    geom_ribbon(
      data = dat.ci,
      aes(x = x, ymin = lower, ymax = upper),
      fill = "steelblue",
      alpha = 0.2
    ) + ggtitle(capture.output(obj$ci))
} 

```


## A function for plotting 95% CI bands

```{r fig.align='center'}
roc_with_ci(roc2) %>% 
  gf_labs(title = "s100b ROC curve with bootstrap 95% confidence bands")

```


## Exercise

Use the `aSAH` dataset in package `pROC` to analize the accuracy of `wfns` (World Federation of Neurological Surgeons) to predict poor outcome.

1. Plot the ROC curve of `wfns` along with those of `ndka` and `s100b`. 

1. Compute AUC ROC and CI for all three markers, using appropriate method: DeLong's or bootstrap?

1. Compare the AUC ROC of `wfns` with the other two markers using appropriate method: DeLong's or bootstrap?

1. Plot the optimal threshold (maximizing the Youden index) in the ROC curve.

`wfns` is a factor. Start by converting to a numeric variable with `as.numeric()`:

```{r eval = FALSE, echo=TRUE}
library(pROC); data(aSAH) 
aSAH$wfns <- as.numeric(aSAH$wfns)
```



