---
title: "General Linear Model (2)"
subtitle: "Assessing model assumptions"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      error = TRUE,
                      comment = NA)

library(tidyverse)
library(mosaic)
data(Births78)

d <- Births78 %>% 
  select(-year) %>% 
  mutate(day = ifelse(wday %in% c("Sun", "Sat"),                  
                      as.character(wday),                      
                      "Work"),
         day = factor(day, levels = c("Work", "Sat", "Sun"))) 

rm(Births78)



```



## Assessing model assumptions

Assumptions:

1. Residuals are normal

2. Variance of residuals is constant


Assessment:

1. QQ plot

2. Plot of Residuals vs Predicted

</br>

In practice:  care with *influential* outliers!


## Residual plots (with base R plotting system)


```{r fig.height=4, fig.width=10}

mod_dd <- lm(births ~ day + poly(day_of_year,5), data = d)
par(mfrow = c(1,2), mar = c(4,8,2,2))    # two plots in a slide, margins
plot(mod_dd, 1)                          # residuals vs fitted
plot(mod_dd, 2)                          # QQ plot

```

## Holidays

See [US holidays in 1978](https://www.timeanddate.com/calendar/?year=1978&country=1)

Define new variable `holiday`  
(only for working days to avoid double discounting if weekend!)

```{r}
h <- as.Date(c("1978-01-02", "1978-02-20", "1978-05-29", "1978-07-04", 
               "1978-09-04", "1978-11-11", "1978-11-23", "1978-12-25"))

d <- d %>% mutate(holiday = ifelse(date %in% h & day == "Work",
                    "Yes", 
                    "No"))

```

## Adding holidays to the model {.smaller}

```{r}
mod_ddh <- lm(births ~ day + holiday + poly(day_of_year,5) , data = d); summary(mod_ddh)

```


## Reassessing assumptions

```{r fig.height=4, fig.width=10}
par(mfrow = c(1,2), mar = c(4,8,2,2))
plot(mod_ddh, 1)
plot(mod_ddh, 2)

```

## Inspecting outliers

```{r}

d %>% 
  mutate(residual = resid(mod_ddh)) %>%       # save residuals to d
  slice(51, 353, 360) %>%                      # filters by row number
  select(date, wday, holiday, residual)

```

1978-02-20: $\quad$ Presidents' Day, a public holiday in most US states. Nonetheless, many businesses and stores are open.

1978-12-26: $\quad$ Day after Christmas. 

1978-12-19: $\quad$ ??


## R-squared for the final model

In the plot, `births` and `predicted` are deviations from their means

```{r echo = FALSE, fig.align='center', fig.height=3}

list(residuals = resid(mod_ddh),
     predicted = predict(mod_ddh) - mean(d$births),
     births = d$births - mean(d$births)) -> k

par(mar = c(2,5,3,1), mfrow=c(1,1))
boxplot(k, border = c("red", "blue", "black"), 
        horizontal = TRUE, las = 1)
title(paste("R-squared:",
                round(summary(mod_ddh)$r.squared,3)))



```

</br>

The model explains `r round(100*summary(mod_ddh)$r.squared,1)` % of the total variation of `births`.  
Not bad!

## Exercise

1. Fit the previous model (`day`, `holiday` and `day_of_year` with 5-degree polynomials).

2. Fit a new model by adding the `day_of_month` to the previous model. 

3. Compare the two models by looking at:

    - R-square  
    
    - Adjusted R-squared (which penalizes model complexity)

4. Are the two models nested? If yes, compare them with `anova()`. Does the addition of `day_of_month` improve the fit?


