---
title: "General Linear Model (3)"
subtitle: "Confounding and interaction"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, warning = FALSE,  message = FALSE,
                      comment = NA, fig.align='center')


```


## Confounding

General interpretation of model coefficients:  
Effect of the corresponding $x$ on $y$, **given all other predictors**

- Coefficients will generally change if we change the model  
(i.e., eliminate, or add, predictors)
- This provides a way to assess **confounding**:
    - Model 1: $\quad y = \beta_0 + \beta_1 x_1 + \epsilon$  
    - Model 2: $\quad y = \beta_0 + \beta_1 x_1 + \beta_2 x_2 + \epsilon$
    
\

$x_2$ is a **counfounder** for $x_1$, if the effect of $x_1$ (that is, $\beta_1$)  
is different in Model 1 and Model 2



## Confounding example (simulated data)

- Suppose an observational study comparing lipid lowering drugs A and B    
- Colesterol is measured pre- and pos-treament  

```{r conf, echo=F, fig.width=10, fig.height=4}

set.seed(100)
x = round(300 + 30*rnorm(100))
k = as.numeric(cut(x,quantile(x)+.5,labels=1:4))
temp=rbinom(100,1,0.5)
t = ifelse(k<2,1,ifelse(k>3,0,temp))
y = round(8.5 + 0.95*x + 10*rnorm(100))


library(RColorBrewer)
myColors <- brewer.pal(6, "Set1")

par(mfrow=c(1,2), mar=c(4,4,2,2))
treatment <- factor(t, levels=0:1, labels=LETTERS[1:2])
boxplot(y ~ treatment, col=myColors, ylab="Pos-treatment cholesterol (mg/dl)")

color = myColors[t+1]
plot(y ~ x, col=color, las=1,  
     ylab="Pos-treatment cholesterol (mg/dl)", 
     xlab="Pre-treatment cholesterol (mg/dl)")
legend(x="topleft",c("drug A","drug B"), pch=1,col=myColors)
rug(y[t==1], side=2, col=myColors[2])
rug(y[t==0], side=4, col=myColors[1])
```



## Modelling pos-treatment values (y)  {.smaller}

- A first model (m1) including only **treatment** as predictor (A is the  reference level)
- A second model (m2) including both **treatment** and **pre-treatment values (x)**  
- In both models, $\beta_{treatmentB}$ is the expected **difference** betwen B and A (the **reference** level)

```{r conf2, echo=TRUE}

m1 <-  lm(y ~ treatment); print(summary(m1)$coeff,digits=2) 

m2 <-  lm(y ~ x + treatment); print(summary(m2)$coeff,digits=2)

```
  
- Note that treatment effect is no longer significant
- Note how small is now the estimated treatment effect


## The two models, grapically

In both plots, the treatment effect is the *vertical* distance between lines (A-B):


$\qquad \qquad$ **m1**: quite obviously, `r round(coef(m1)["treatmentB"])` mg/dl $\qquad \qquad  \qquad$  **m2**: *given* $x$, `r round(coef(m2)["treatmentB"])` mg/dl


```{r echo = FALSE, fig.width=10, fig.height=4}

par(mfrow=c(1,2), mar=c(4,4,2,2))
plot(y ~ jitter(t,1/2), col=myColors[t+1], 
     ylab="Pos-treatment cholesterol (mg/dl)",
     xlab="Treatment",
     xlim = c(-0.5,1.5),
     axes=FALSE,
     frame.plot=TRUE)
axis(2)
axis(1, at=0:1, labels=c("A", "B"))
abline(h = coef(m1)[1], col = myColors[1])
abline(h = coef(m1)[1]+coef(m1)[2], col = myColors[2])
title(as.character(m1$call)[2])

plot(y ~ x, col=color, las=1,  
     ylab="Pos-treatment cholesterol (mg/dl)", 
     xlab="Pre-treatment cholesterol (mg/dl)")
legend(x="topleft",c("drug A","drug B"), pch=1,col=myColors)
rug(y[t==1], side=2, col=myColors[2])
rug(y[t==0], side=4, col=myColors[1])
abline(a = coef(m2)[1], b = coef(m2)[2], col = myColors[1])
abline(a = coef(m2)[1]+coef(m2)[3], b =coef(m2)[2] , col = myColors[2])
title(as.character(m2$call)[2])
```


## Interaction

Interaction between two predictors $x_1$ and $x_2$:

- the effect of $x_1$ is not constant, but depends on $x_2$ 

- the effect of $x_2$ is not constant, but depends on $x_1$


</br>

If this is the case, we say that

- $x_2$ is an effect modifier for $x_1$

- $x_1$ is an effect modifier for $x_2$


## Interaction example {.smaller}

- Clinical trial on a lipid lowering drug vs placebo; blood colesterol is measured pre- and pos-treatment  
- Note in this case pre-treatment values are similar in both treatment groups

```{r inter_example, echo=F, fig.width=5}
set.seed(100)
x <-  round(300 + 30*rnorm(100))
t <-  rep(0:1,each=50)
y <-  round(8.5 + 0.95*x + 80*t -0.36*x*t + 10*rnorm(100))
color <-  myColors[t+1]

par(mfrow=c(1,1), mar=c(4,4,1,2))
plot(x,y, col=color, las=1, ylab="Pos-treatment", xlab="Pre-treatment")
legend(x="topleft",c("Pacebo","Drug"), pch=1, col=myColors[1:2])

```



## Fitting models with/without interaction 

```{r inter_model, echo=TRUE, fig.width=5}
m1 <-  lm(y ~ x + t); print(summary(m1)$coeff, digits=2)          # without interaction
m2 <-  lm(y ~ x + t + x:t); print(summary(m2)$coeff, digits=2)    # with interaction 

```



## The two models, grapically

In both cases we fit two lines (one for each treatment), but:

$\qquad$ different intercept, same slope $\qquad$ different intercept, different slope

```{r inter_plots, echo=F, fig.width=10}

par(mfrow=c(1,2), mar=c(4,4,4,2))
plot(x,y, col=color, las=1, ylab="Pos-treatment", xlab="Pre-treatment")
abline(m1$coefficients[1:2])
abline(m1$coefficients[1]+ m1$coefficients[3], m1$coefficients[2])
legend(x="topleft",c("Pacebo","Drug"), pch=1, col=myColors[1:2])
title("m1: without interaction")

plot(x,y, col=color, las=1, ylab="Pos-treatment", xlab="Pre-treatment")
abline(m2$coefficients[1:2])
abline(m2$coefficients[1]+ m2$coefficients[3], m2$coefficients[2]+m2$coefficients[4])
legend(x="topleft",c("Pacebo","Drug"), pch=1, col=myColors[1:2])
title("m2: with interaction")

```



## Exercise 1

Use this code to read and prepare the `GLMsData::lungcap` dataset on lung capacity and smoking in 654 youth (see `?GLMsData::lungcap`).

```{r}
library(tidyverse)
library(GLMsData)
data(lungcap)                                            # get the dataset

d <- lungcap %>% 
  janitor::clean_names() %>%
  mutate(height = round(2.54 * ht),                       # converts from inches to cm
         gender = as.factor(gender),                      # creates factor for gender
         smoke = factor(smoke,                            # creates factor for smoke
                        levels = 0:1,
                        labels = c("no", "yes"))) %>% 
  select(-ht)                                             # drops ht

```

## Exercise 1 (cont.)

1. Get a summary of the dataset variables

1. Subset non-smokers into dataframe `dns`

1. Fit a model for FEV with `height` and `gender` as  explanatory variables and:
  - get a summary of the model
  - interpret the model coefficients
  - plot the `fev` against `height` using color for gender,and overlay predicted values.
  - assess model assumptions graphically. Is there any non-linearity?  If yes, does a polynomial improve the fit?  
  

## Exercise 2
  
1. Using all cases in dataframe `d`, produce boxplots of:
    - `fev` by `smoke`, and 
    - `age` by `smoke`. 

1. Fit a model for `fev` with `smoke` as predictor; interpret the model coefficients.

1. Fit a second model adding `age` as predictor; do you think `age` is a confounder for `smoke`?
