---
title: "Logistic Regression model (1)"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)

```

## Burns

`burn1000` in package `aplore3`:  1000 cases of burn injuries

```{r}
library(aplore3)
d <- burn1000         # just to have shorter name for the dataframe
head(d, 4)
```

- `tbsa`: $\qquad$ Total burn surface area (0 - 100%)
- `inh_inj`: $\quad$ Burn involved inhalation injury 
- `flame`: $\qquad$ Flame involved in burn injury 


## A summary of the data

```{r}
library(tidyverse)
d %>% 
  select(-id, -facility) %>% 
  summary()
```


## Modeling probabilities - Logistic function {.smaller}

Outcome coded as 1 for "Dead", 0 for "Alive". Does the *probability* of dying ($\pi$) increase with `tbsa` ($x$)?

</br>

```{r, echo=FALSE, fig.height=3, fig.width=10, fig.align='center'}
outcome <- ifelse(d$death == "Dead", 1, 0)

m <- lm(outcome ~ tbsa, data = d)
d$pred <- predict(m)

library(ggformula)
gf_jitter(outcome ~ tbsa, data =d, alpha = 0.2, width=0, height=0.05,
          title="Linear function") %>%
  gf_line(pred ~ tbsa, color = "red", lwd=1) %>%
  gf_abline(slope=0, intercept = 1, lty=2) %>%
  gf_abline(slope=0, intercept = 0, lty=2) %>%
  gf_lims(y = c(-0.10, 1.10)) +
  scale_y_continuous(breaks = seq(0,1,.2)) -> p1


m <- glm(outcome ~ tbsa, family = binomial, data = d)
d$pred <- predict(m,  type = "response")

gf_jitter(outcome ~ tbsa, data =d, alpha = 0.2, width=0, height=0.05,
          title="Logistic function") %>%
  gf_line(pred ~ tbsa, color = "red", lwd=1) %>%
  gf_abline(slope=0, intercept = 1, lty=2) %>%
  gf_abline(slope=0, intercept = 0, lty=2) %>%
  gf_lims(y = c(-0.10, 1.10)) +
  scale_y_continuous(breaks = seq(0,1,.2)) -> p2

library(gridExtra)
grid.arrange(p1 + theme(plot.margin=margin(2,40,2,40)),
             p2 + theme(plot.margin=margin(2,40,2,40)),
             nrow=1, ncol=2)

```

</br> 

<center>
$\pi = \beta_0 + \beta_1 x \qquad \qquad \qquad  \qquad  \pi = \frac{e^{\beta_0 + \beta_1 x}}{1 \enspace + \enspace e^{\beta_0 + \beta_1 x}}$ 
</center>

## The logit transformation 

$\pi = \frac{e^{\beta_0 + \beta_1 x}}{1 \enspace + \enspace e^{\beta_0 + \beta_1 x}} = \frac{e^{LP}}{1 \enspace + \enspace e^{LP}} \qquad$ where $\quad LP = \beta_0 + \beta_1 x$

$\frac{1}{\pi} = \frac{1 \enspace + \enspace e^{LP}}{e^{LP}} = \frac{1}{e^{LP}} + 1$

$\frac{1}{\pi} -1 = \frac{1}{e^{LP}}$

$\frac{1-\pi}{\pi} = \frac{1}{e^{LP}}$

$\frac{\pi}{1-\pi} = e^{LP} \quad \implies \quad ln(\frac{\pi}{1-\pi}) = \beta_0 + \beta_1 x$

To note:

- $ln(\frac{\pi}{1-\pi})$ is the *log-odds*, or the *logit* of $\pi$

- The logit of $\pi$ is linearly related to x

## A first model based on `tbsa` {.smaller}

Use function `glm()` with option `familiy = binomial` !!  
You will be modeling the probability of the *second* level of `death` !!

```{r}
mod_s <- glm(death ~ tbsa, data = d, family=binomial)         # do not forget!  family=binomial
summary(mod_s)

```

## Interpreting coefficients of quantitative vars

<center>
$ln(\frac{\pi}{1-\pi}) = \beta_0 + \beta_1 x$  
</center>
</br>

$\beta_0 =$ log-odds when $x=0$  

$\beta_1 =$ change in log-odds per unit increment in $x$  
- $x=k: \quad \qquad \beta_0 + \beta_1 k$  
- $x=k+1: \quad \beta_0 + \beta_1 (k+1) = \beta_0 + \beta_1 k + \beta_1$  

So that,   $\qquad \beta_1 = ln(\frac{\pi|x=k+1}{1-\pi|x=k+1}) - ln(\frac{\pi|x=k}{1-\pi|x=k}) = ln \left( \frac{\frac{\pi|x=k+1}{1-\pi|x=k+1}}{\frac{\pi|x=k}{1-\pi|x=k}} \right)$  
Therefore, $\qquad e^{\beta_1} = \frac{\frac{\pi|x=k+1}{1-\pi|x=k+1}}{\frac{\pi|x=k}{1-\pi|x=k}} \qquad$ is the OR for a unit increment in $x$

## Exponentiate coefficients to interpret

```{r}
coef(mod_s); exp(coef(mod_s)) 
```

The odds when `tbsa`=0:   $\quad \text{O}_0 =$ `r round(exp(coef(mod_s))[1],2)`  
(meaningless in this case, since `tbsa`=0 means no burn!)

The OR for a unit increase in `tbsa`:  $\quad \text{OR}_1 =$ `r round(exp(coef(mod_s))[2],2)` 

In general, the OR for a $k$ units increase is $(\text{OR}_1)^k$, therefore 

The OR for a 10 units increase in `tbsa`: $\quad$ (`r round(exp(coef(mod_s))[2],2)`)$^{10}$ = `r round(exp(coef(mod_s))[2]^10,2)`

## Exponentiate the CI of coefficients also

```{r}
exp(coef(mod_s))           # exponential of coefficients
exp(confint(mod_s))        # and their 95% CI
```

```{r echo=FALSE}
exp(coef(mod_s))[2] -> beta
exp(confint(mod_s))[2,] -> beta_ci
```

The OR for a unit increase in `tbsa` is $\quad$ `r round(beta,2)`  [`r  paste(round(beta_ci,2), collapse=", ")`]

and raising these values to 10...

The OR for a 10 units increase in `tbsa` is $\quad$ `r round(beta^10,2)`  [`r paste(round(beta_ci^10,2), collapse=", ")`]

## A second model, based on `inh_inj` {.smaller}

```{r}
mod_i <- glm(death ~ inh_inj, data = d, family=binomial)  
summary(mod_i)
```

## Interpreting (exponentiated) coefficients

If explanatory variable is a factor, the *exponentials* of coefficients are:

- `Intercept`: the odds of dead for the reference level (no inhalation injury)

- `inh_injYes`: the OR of dead (Inhalation injury / no inhalation injury)

```{r}
exp(coef(mod_i))

library(mosaic); tally(death ~ inh_inj, data=d)

```

## Exercise

1. Read the [Titanic data](https://raw.githubusercontent.com/acobos/Datasets/master/Mysterious_event.txt) with the following code:

    ```{r eval = FALSE}
    d <- rio::import("https://raw.githubusercontent.com/acobos/Datasets/master/Mysterious_event.txt",
                 stringsAsFactors  = TRUE)
    ```

    + What are the levels of `Survival`? Which level will be modelled in a logistic regression analysis using `glm(..., family = binomial)`?
    
    + Fit a logistic regression model (`m1`) for `Survival` with only `Class` as predictor, and get the model summary.
    
    + Based on the summary, what class has the lowest survival probability?
    
    + What is the OR of survival for class S compared to class 1?

    + What is the OR of survival for class 1 compared to class S?  

