---
title: "Logistic Regression model (2)"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)



```

## Burns data

```{r}
library(tidyverse)
library(aplore3)
d <- burn1000         # just to have shorter name for the dataframe
head(d)

mod_s <- glm(death ~ tbsa, data = d, family=binomial)
mod_i <- glm(death ~ inh_inj, data = d, family=binomial)  
```


## A third model with two predictors {.smaller}

```{r}
mod_si <- glm(death ~ inh_inj + tbsa, data = d, family=binomial)   
summary(mod_si)
```

## Comparing nested models: Likelihood Ratio Test

Use `anova()` with option `test="LRT"` 

```{r}
anova(mod_s, mod_si, test = "LRT")
```

Deviance is a measure of (lack of) fit: the lower, the better

AIC is similar, but penalizing model complexity: the lower the better

## The effect of inhalation injury {.smaller}

```{r fig.height=2.5, fig.align='center'}
exp(coef(mod_i)); exp(coef(mod_si))

library(ggformula)
gf_boxplot(tbsa ~ inh_inj, data=d) + coord_flip()
```

## Predicted values

```{r}
predict(mod_si) %>% head()       # gives the logits!
```

\
\

to get predicted probabilities use argument `type="response"` 

```{r}
predict(mod_si,  type = "response") %>% head()
```



## Assumptions 

1. Large sample: 10-25 events and non-events per *parameter* (not per variable!) 

2. Linearity on the logit scale (numeric predictors)

3. No *influential* outliers

</br>

(1) In the burns data, 150 deaths and 850 alive, so:

- Models with > 150/10 = 15 parameters are unreliable

- Models with $\le$ 150/25 = 6 parameters are OK

- Models with 7-15 parameters, care

## Events per parameter {.smaller}

How many parameters are estimated in this model?

```{r}
mod <- glm(death ~ (inh_inj + flame) * (tbsa * age) + gender + race , data=d, family = binomial)
round(summary(mod)$coef,4)
```



## A model with six variables {.smaller}

```{r}
m6v <- glm(death ~ tbsa + inh_inj + flame + age + gender + race , 
           data=d, family = binomial)
summary(m6v)

```


## Linearity on the logit scale

```{r fig.align='center', fig.width=10, fig.height=3.5}

library(patchwork)
plot1 <- gf_point(predict(m6v) ~ tbsa, data = d) %>% gf_smooth(method = "loess")
plot2 <- gf_point(predict(m6v) ~ age, data = d) %>% gf_smooth(method = "loess")
plot1 + plot2

```

## Influence measures 

Cook's distance, standardized residuals and leverages

```{r fig.align='center', fig.height=3.5}
par(mfrow = c(1,2))
plot(m6v,4)
plot(m6v,5)
```


## Exercise

1. Read the [Titanic data](https://raw.githubusercontent.com/acobos/Datasets/master/Mysterious_event.txt) as in the previous exercise and fit a model (`m2`) for `Survival` with `Class`, `Age` and `Sex` as predictors:

    + What is the class with the lowest survival probability, given sex and age?  

    + What is the OR of survival for females compared to males, for any given combination of class and age?  

    + Use a Likelihood Ratio Test to compare model `m2` to model `m1` fitted in the previous exercise (with only `Class` as predictor).
    
    + Is the fit of `m2` significantly better than that of `m1`? 