---
title: "Survival analysis"
subtitle: "with packages `survival` and `survminer`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)
```

## Survival analysis

Variable: $\quad$ time to an event (T), e.g., death, disease progresion, etc.

Problem: $\quad$ _censored_ observations (right-censoring)

- loss to follow-up 

- no event by end of fixed study period

Goals:  

- describe survival as a function of time: $\quad$ _Kaplan-Meier_ method

- compare survival in different sets of patients: $\quad$  _log-rank_ test

- model survival as a function of covariates (age, sex, treatment, etc.):  
$\quad$ _Cox proportional hazards_ model


## Right censoring {.smaller}

```{r echo=FALSE, fig.align='center', fig.width=12, fig.height=4}
library(survival)
data(ovarian)

library(dplyr)
ovarian %>%
  arrange(futime, fustat) %>%
  mutate(status = factor(fustat,
                         levels = 0:1,
                         labels = c("death", "censored")),
         pid = 1:nrow(ovarian)) %>%
  select(pid, days = futime, status) -> d


library(ggformula)

gf_segment(0 + days ~ pid + pid, data = d) %>%
  gf_point(days ~ pid, shape = 16, data = d[d$status == "death",],
           show.legend = TRUE) %>%
  gf_labs(y = "Time (days)", x = "Patients", title="Dots are events") %>%
  gf_hline(yintercept = 365 * 1:3, lty = 2) +
  scale_x_reverse(lim=c(26,1), breaks = 1:26) +
  coord_flip() -> plot

library(gridExtra)
library(grid)

tableGrob(head(d,11), rows = NULL) -> dt
            
grid.arrange(plot, dt, ncol=2)

  
```


How to represent censoring in a single variable?

How to estimate survival at different times in the presence of censoring?


## Survival and hazard functions {.smaller}

__survival function__: probability of surviving _at least_ until time $t$ $\qquad \small{S(t) = \text{Prob}(T > t) = 1- F(t)}$

__hazard function__: instantaneous death / event _rate_ at time $t$  $\qquad h(t) = \frac{f(t)}{S(t)} = - \frac{d \log S(t)}{d t}$

__cumulative hazard function__:  accumulated risk until time $t$ $\quad \quad \quad \small{H(t) = - \log S(t)}$


```{r echo =FALSE, fig.align='center', fig.height=3, fig.width=10}

library(survival)
data(lung)

fit <- survfit(Surv(time, status) ~ 1, data = lung)

par(mar=c(6,6,1,2), mfrow=c(1,3))

# S(t)
plot(fit, conf.int = FALSE, las = 1, 
     xlab = "time (days)", ylab = "Survival probability")

# h(t)
library(bshazard)
h <- bshazard(Surv(time, status) ~ 1, data = lung, verbose=FALSE)

plot(h, conf.int = FALSE, las=1,
     xlab = "time (days)",  yaxt="n")
axis(2, las=1, cex.axis=0.7)

# H(t)
plot(fit$time, -log(fit$surv), type="n", las=1,
     xlab = "time (days)", ylab = "Cumulative hazard")
lines(fit$time, -log(fit$surv))



```

$h(t)$ is conditional on not having experienced the event before t. It's _not_ a probability.  
$H(t)$ does not have an easy interpretation.



## Advanced lung cancer

Survival in `r nrow(survival::lung)` patients with advanced lung cancer from the North Central Cancer Treatment Group (Journal of Clinical Oncology. 12(3):601-7, 1994).  
See `?lung` for more details.

```{r}

library(survival)
head(lung, 4)

```

<font size="4">
- time: survival time in days  
- status: censoring status 1=censored, 2=dead  
- sex: Male=1 Female=2  
</font>


## Survival objects {.smaller}

The `Surv()` function creates suvival objects combining time and censoring/event information. You need to specify:

- `time`  (the follow-up time) 

- `event` (see `?Surv` for allowed event codings)

The result is a value for each subject: the time followed by + if censored


```{r}

surv_obj <- Surv(time = lung$time, event = lung$status)        # creates a survival object
                                                               
head(surv_obj, 20)                                             # prints first 20 values                    

```



## Kaplan-Meier (KM) curve {.smaller}

The `Survfit` function estimates survival using KM method. You need to specify a _formula_ with a survival object (before the tilde), and either 1 (for a single curve) or a factor (for a curve for each level).

```{r fig.align='center', fig.height=3, fig.width=5}

km <- survfit(Surv(time = time, event = status) ~ 1, data = lung)           # KM estimation
library(survminer)
ggsurvplot(km, xlab = "Days", legend = "none",                              # plots  survfit objects
           censor=FALSE, surv.median.line = "hv")     

```

See `?ggsurvplot` for more customization options

## KM estimates {.smaller}

- Get KM survival estimates at specific times using function `summary()` on a `survfit` object.
- Get median survival time by printing the `survfit` object  

```{r}

summary(km, time = c(365, 730))     # survival estimates at specified times
km                                  # median survival time

```

## KM estimates {.smaller}

- Get other quantiles using the `quantile()` function on the `survfit` object  

```{r}

quantile(km)

```



## Log-rank test {.smaller}

To compare survival curves with this test, use function `survdiff()` on a formula, with the factor defining groups after the tilde.

```{r}
lung <- lung %>% mutate(sex = factor(sex, levels=1:2, labels = c("male", "female")))
survdiff(Surv(time, status) ~ sex, data = lung)      # note formula

```

This test can be used if:

- Groups are independent (different subjects)

- With similar, non-informative censoring pattern, and

- Proportional hazards $\implies$ curves do not cross

## Survival curves with log-rank test {.smaller}

```{r fig.align='center', fig.height=4, fig.width=6}

ggsurvplot(survfit(Surv(time, status) ~ sex, data = lung),
           pval = TRUE,                                     # plots the p-value
           pval.method = TRUE)                              # plots the method (test)

```




## Exercise {.smaller}

Using the `ovarian` dataset in package `survival` (see `?ovarian`):

1. Define factors for `rx (1="A", 2="B")`, `resid.ds (1="no", 2="yes")` and `ecog.ps (1="good", 2="bad")`, and factor to dichotomize age, with labels "old" (if age >=50), or "young" (otherwise).

1. Create a survival object from `futime` and `fustat`, and print it.

1. Use this object to fit a KM curve for each treatment group, and plot them using option `risk.table = TRUE` in the `ggsurvplot()` function call.

1. What is the median survival time for each treatment?

1. What are the survival estimates at 1 and 2 years?

1. Compare the two curves with a log-rank test.

1. Examine the predictive value of the residual disease by plotting a survival curve for each residual disease group, showing the result of the log-rank test.




