---
title: "Cox model"
subtitle: "with packages `survival` and `survminer`"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)

```

## The Cox _proportional hazards_ model (CPHM) {.smaller}


<center>

$h(t) \quad = \quad h_0(t) \quad e\ ^{\ \beta_1 x_1 + \ \beta_2 x_2 + ...+ \ \beta_k x_k} \quad = \quad  h_0(t) \quad e\ ^{\beta_1 x_1} \ e\ ^{\beta_2 x_2} \ ... \ e\ ^{\beta_k x_k}$ 

</center>

Models the _hazard function_ $h(t) = \frac{f(t)}{S(t)} = - \frac{d \log S(t)}{d t}$, the instantaneous _hazard_ of event at time $t$

$h_0(t)$ is the baseline hazard (i.e., when all $x$'s are 0)

$e\ ^{\beta}$ are _hazard ratios_; suppose $x_1$ is an indicator (i.e., 0 or 1); then  

- $h(t|x_1=1) \quad = \quad h_0(t) \quad e\ ^{\beta_1} \quad e\ ^{\beta_2 x_2} \quad ... \quad e\ ^{\beta_k x_k}$

- $h(t|x_1=0) \quad = \quad h_0(t) \quad 1 \qquad e\ ^{\beta_2 x_2} \quad ... \quad e\ ^{\beta_k x_k}$

So that  

- $\frac{h(t|x_1=1)}{h(t|x_1=0)} = e\ ^{\beta_1}$

So, it is implicitely assumed that the _hazard ratio_ for _any time t_ is a constant, which does not depend on $t$


## Data prep

```{r}
library(tidyverse)
library(survival)
d <- lung %>% 
  mutate(status = factor(status, 
                         levels = 1:2, 
                         labels = c("censored", "dead")),
         sex = factor(sex, 
                      levels = 1:2, 
                      labels = c("male", "female")))
head(d)
```



## Fitting a CPH model {.smaller}

Function `coxph()` fits CPH models

```{r}

fit <- coxph(Surv(time, status == "dead") ~ sex, data = d)
fit

```

\

Assumptions:

- Non-informative censoring

- Proportional hazards 

## Hazard Ratios (HR) {.smaller}

Exponentiate the model coefficients (and CI) to get HR (and CI)

```{r}
coef(fit)           # gets model coefficients
exp(coef(fit))      # gets the HR 
exp(confint(fit))   # gets the 95% CI for HR
```


## A model with three variables {.smaller}

```{r}

fit <- coxph(Surv(time, status == "dead") ~ age + sex + ph.ecog, data = d)
fit

exp(coef(fit))                    # to get hazard ratios (HR)

```


## Plotting HR {.smaller}

```{r fig.align='center'}

survminer::ggforest(fit, data = d, fontsize = 1)            

```


## Assessing the PH assumption {.smaller}

A test for deviation from proportional hazards can be done with function `cox.zph` 

```{r}

cox.zph(fit)

```

A significant p-value means evidence of deviation from proportional hazards.

Fore more thorough ways of assessing assumptions of the Cox PH model, see [this](https://www.r-bloggers.com/cox-model-assumptions/)

When PH do not hold, [Accelerated Failure Time (AFT) models](https://pubmed.ncbi.nlm.nih.gov/17080754/) may be used.

## Exercise 

Prepare the `ovarian`  data as in the previous exercise.

1. Fit a Cox PH model with treatment group, residual disease, age group, and ECOG performance status as covariates.

1. Plot the HR from this model fit

1. Verify the PH assumption in this model fit


