---
title: "Quantitative variables"
subtitle: "Paired samples"
author: "Albert Cobos"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning=FALSE, 
                      error = TRUE,
                      comment = NA,
                      message = FALSE)

# library(pander)
```

## Weight and smoking cessation

```{r echo=TRUE}
d <- rio::import("data/smoking_cessation.txt")
head(d)
```

Is there a change in weight after smoking cessation?

We want to compare weight `pre`- and `pos`-smoking cessation


## Tests to compare two paired distributions {.smaller}

If distribution of *individual differences* is about normal$^1$: $\quad$ parametric test

- paired, or one-sample *t*-test

Otherwise: $\quad$ Distribution-free ("non-parametric") tests

- Wilcoxon signed ranks test

</br>

$^1$ In practice, parametric test works well if:  
- distribution of *individual differences* is (approximately) symmetrical  
- no *influential* outliers

which can be inspected in a boxplot.

\ 

Paired tests are based on the *individual differences*.

## Paired data and individual differences

The mean of individual differences = difference of means

```{r}
library(tidyverse)
d <- mutate(d, change = pos - pre)   # compute individual differences
head(d, 3)
sapply(d, mean)                      # to get the mean of all variables in d
```

``r paste0(mean(d$pos), " - ", mean(d$pre), " = ", mean(d$pos) - mean(d$pre))``


## Boxplot (with mean overlaid)

```{r echo = TRUE, fig.height=1, fig.align='center'}
library(ggformula)
gf_boxplot(change ~ "", data = d, xlab = "", ylab = "change = pos - pre") %>% 
  gf_summary(fun = mean, geom = "point", color = "red") %>% 
  gf_refine(coord_flip())

```

In this case:

- distribution is  (approximately) symmetrical

- a single outlier in smokers, but not influential (mean $\approx$ median)

Parametric test is OK

## Paired, or one-sample *t*-test {.smaller}

```{r}
t.test(d$change)                               # one-sample t-test on the differences
t.test(d$pos, d$pre, paired = TRUE)            # paired t-test on pos vs pre
```

## When conditions for the *t*-test are not met

Murder rates (per 100000 inhabitants) by gender in 198 countries

```{r }
load("data/murder_rates.RData")
head(d)
```

</br>

source:  https://www.gapminder.org/data/


## Differences in rates (men - women)

```{r}
d$difference <- d$men - d$women    # compute country differences to plot them
```


```{r echo=FALSE, fig.height=1.5}
gf_boxplot(difference ~ "", data=d, xlab="", ylab="difference (men - women)") +
  stat_summary(fun.y=mean, geom="point", color="red") +
  coord_flip()
```

In this case:

- distribution is  clearly assymmetrical

- several outliers, with influence (mean > median)

Distribution-free test (Wilcoxon signed ranks test)

## Wilcoxon signed ranks test

```{r}
wilcox.test(d$difference)
```

## Exercise

Read [this file](https://raw.githubusercontent.com/acobos/Datasets/master/Life_Expectancy_with_continent.csv). This should be the first 6 lines of the resulting dataframe:

```{r echo = FALSE}
d <- rio::import("https://raw.githubusercontent.com/acobos/Datasets/master/Life_Expectancy_with_continent.csv")
head(d)
```

Try to answer these questions:

1. Is life expectancy the same for men and women?

2. Provide some estimate of the difference

